Katana: Towards Patching as a Part of the ELF Toolchain

### 2010-03-27 paper1/hotpatch/abstract.tex
many grammar fixes on advice from a retired english teacher

\begin{abstract}
Despite advances in software modularity, security, and reliability,
offline patching remains the predominant form of updating or
protecting commodity software.  Unfortunately, the mechanics of {\it
  hot patching} (the process of upgrading a program while it executes)
remain understudied, even though such a capability offers practical
benefits for both consumer and mission-critical systems.
%
A reliable hot patching procedure would serve particularly well by
reducing the downtime necessary for critical functionality or security
upgrades.  Yet, hot patching also carries the risk -- real or
perceived -- of leaving the system in an inconsistent state, which
leads many owners to forego its benefits as too risky.  
%
%MEL:
%entity -- an
%application binary, library or even the kernel, while it is still in
%execution. 
%SB:
% It offers enormous benefits as it helps in mitigating
% process downtime and contributes to overall system reliability and
% availability.
%
%It offers great practical benefits for mission-critical systems, by
%reducing downtime for upgrades essential for maintaining their {\em
%  reliability} and {\em availability}. 
%
In this paper, we propose a novel method for hot patching ELF binaries
that supports {\em (a)} synchronized global data and code updates and
{\em (b)} reasoning about the results of applying the hot patch.  We
propose a format, which we call a {\it Patch Object}, for encoding patches as
a special type of ELF relocatable object file.  Our tool, {\em
  Katana}, automatically creates these patch objects as a by-product of the
standard source build process.  Katana also allows an end-user to
apply the Patch Objects to a running process.  In essence, our method
can be viewed as an extension of the Application Binary Interface
({\it ABI}), and we argue for its inclusion in future ABI standards.

>>>>

\begin{abstract}
Despite advances in software modularity, security, and reliability,
offline patching remains the predominant form of updating or
protecting commodity software.  Unfortunately, the mechanics of {\it
  hot patching} (the process of upgrading a program while it executes)
remain understudied, even though such a capability offers practical
benefits for both consumer and mission-critical systems.
%
A reliable hot patching procedure would serve particularly well by
reducing the downtime necessary for critical functionality or security
upgrades.  Yet, hot patching also carries the risk -- real or
perceived -- of leaving the system in an inconsistent state, which
leads many owners to forgo its benefits as too risky.  
%
%MEL:
%entity -- an
%application binary, library or even the kernel, while it is still in
%execution. 
%SB:
% It offers enormous benefits as it helps in mitigating
% process downtime and contributes to overall system reliability and
% availability.
%
%It offers great practical benefits for mission-critical systems, by
%reducing downtime for upgrades essential for maintaining their {\em
%  reliability} and {\em availability}. 
%
In this paper, we propose a novel method for hot patching ELF binaries
that supports {\em (a)} synchronized global data and code updates and
{\em (b)} reasoning about the results of applying the hot patch.  We
propose a format, which we call a {\it Patch Object}, for encoding patches as
a special type of ELF relocatable object file.  Our tool, {\em
  Katana}, automatically creates these patch objects as a by-product of the
standard source build process.  Katana also allows an end-user to
apply the Patch Objects to a running process.  In essence, our method
can be viewed as an extension of the Application Binary Interface
({\it ABI}), and we argue for its inclusion in future ABI standards.

### 2010-03-27 paper1/hotpatch/conclusion.tex
many grammar fixes on advice from a retired english teacher

We introduce a method for hot patching: a technique we believe to be a
promising alternative to redundancy, ad hoc self--healing techniques,
``patch and pray,'' or other approaches to dynamic software updates.
Hot patching has the potential for aligning actual practices with
acknowledged ``best practices'' relating to critical security or 
functionality updates.  We hold that one major impediment to hot
patching is the opaque nature of most patches (be it
proprietary or open software), and our method of patching along with the
PO file format are first attempts at providing a basis for informed reasoning
about the structure and implications of a patch.

>>>>

We introduce a method for hot patching: a technique we believe to be a
promising alternative to redundancy, ad hoc self--healing techniques,
``patch and pray,'' or other approaches to dynamic software updates.
Hot patching has the potential for aligning actual practices with
acknowledged ``best practices'' relating to critical security or 
functionality updates.  We hold that one major impediment to hot
patching is the opaque nature of most patches (be it
proprietary or open software), and our method of patching, along with the
PO file format, is a first attempt at providing a basis for informed reasoning
about the structure and implications of a patch.

### 2010-03-27 paper1/hotpatch/design.tex
many grammar fixes on advice from a retired english teacher

In our prototype, a patch is may be generated as follows. The source
directory corresponding to the target to be patched is replicated and
the source code is patched to the version desired. The modified source
tree is then built, and compared with the original source tree at the
object (.o) level. Those object files which have changed between the
modified and original source trees are added to the list of objects
which must be examined for type and code transformations. Future work
will include use of the {\it inotify} mechanism to avoid potentially
expensive recursive directory comparison and provide more precise
notification of changed files.

@@

To dynamically update the running application, Katana needs to patch
both the {\bf code} and the {\bf data} within the process.  It first
creates a {\it patch object} (PO): an ELF file with sections that
indicate the type of patch (code or data), the patch offsets and
lengths within the process address space, patch data, function and
data names, {\it etc.}. The patch object may then be applied to the
target at any time.
%The following sections explain how Katana
%handles both code and data patching.

>>>>

In our prototype, a patch is may be generated as follows. The source
directory corresponding to the target to be patched is replicated, and
the source code is patched to the version desired. The modified source
tree is then built and compared with the original source tree at the
object (.o) level. Those object files that have changed between the
modified and original source trees are added to the list of objects
that must be examined for type and code transformations. Future work
will include use of the {\it inotify} mechanism to avoid potentially
expensive recursive directory comparison and provide more precise
notification of changed files.

@@

To dynamically update the running application, Katana needs to patch
both the {\bf code} and the {\bf data} within the process.  It first
creates a {\it patch object} (PO): an ELF file with sections that
indicate the type of patch (code or data), the patch offsets and
lengths within the process address space, patch data, function and
data names, {\it etc}. The patch object may then be applied to the
target at any time.
%The following sections explain how Katana
%handles both code and data patching.

### 2010-03-27 paper1/hotpatch/discussion.tex
updated paper with more info about safety, limitations of type patching, thoughts on multithreading
many grammar fixes on advice from a retired english teacher

{\bf When to apply the patch.}
%
Dynamically updating a running application requires diligence and
patience. One cannot update the target application without any
knowledge of the program's execution state, by which we mean the
program stack, processor registers {\it etc}.  Even after possessing
this information, the application has to be in what we call a ``safe
state'' for Katana to apply the patch. We characterize a program state
as a {\it safe state} if the following two conditions hold:
\begin{itemize}
\item All activation frames in the program stack belong to functions
  that {\it do not} get updated during code patching. It is easy to
  verify this by comparing each function on the stack to the list of
  upgradeable functions contained within the PO.

\item All activation frames in the program stack belong to functions
  that do not access any global/static symbols identified during {\it
    Data Traversal}, and do not define any local variables of the
  modified types identified during {\it Type Discovery}. Again, since
  we maintain type and variable definitions within the PO, verifying
  this condition is easy.
\end{itemize}

Given the patch object, Katana uses Linux's {\tt ptrace} interface to
temporarily halt the execution of the target process, query the
current execution stack and determine if the application is currently
in a safe state. If so, then Katana applies the code patch followed by
the data patch. We note that it is not possible to apply the code and
data patches at different times since new code likely uses the new
data, and hence postponing data patching to when only the second
condition is satisfied is impractical and unsafe.

Let $A$ denote be the current activation frame on the program stack,
and the notation $(X:Y)$ define all frames from $X$ to $Y$ (on a time
scale, $X$ precedes $Y$) on the stack; so $(1:A)$ defines the current
stack. Now, in case we determine $(1:A)$ to be an unsafe state, we
could repeatedly keep querying the stack until the application reaches
some safe state. However, this is highly inefficient and cumbersome.

@@

Finally we note that even after inserting a breakpoint, the problem of
determining {\it when} the breakpoint will be hit is essentially a
hard one, and so in such cases, Katana can provide no time-bounding
guarantees. Still, this is a cleaner and more efficient approach than
just na\"{i}vely retrying the full update procedure, which is both an
expensive and incomplete solution.
%JMHO:
In cases where the timely (or even eventual) application of a patch
seems unlikely, it would be possible to add support for programmers to
manually specify safe places for patching to occur. For programs built
around a central loop (the vast majority of long-running programs),
the likelihood of safe patching can be expedited by isolating the main
loop into a simple routine unlikely to ever need patching.
%JHMO end

@@

Patching %``randomly'' 
relocated code with our PO files requires
knowledge of the displacements introduced at loading-and-relocation
time. While there is no common ABI standard for saving this
information, conceptually it is no different from saving virtual
addresses of other files' loaded symbols in the {\it Global Offset Table}
(GOT). We note that the names of the constituent object files themselves 
are customarily included in the symbol tables, and that the symbol
table entry format can be easily adopted for storing virtual addresses
of the %``randomly'' 
relocated objects.

Thus, at the cost of small modifications to the OS loader and the 
dynamic linker, we can make the information on the layout of the
``randomly'' relocated executable and libraries available to
our patching process driven by our POs.\footnote{We note that saving 
this information 
%
about the post-relocation layout of the process 
does not weaken ``randomization'', because 
the latter does not assume the attacker's ability to arbitrarily read 
process memory (in which case the addresses of required symbols are
easily found by scanning it %). %
for their code or data patterns), 
but rather breaks hard-coding of these symbols' expected addresses.
}

{\bf Future work.} Katana is a work in progress.
%JMHO:
We have demonstrated code patching (including dynamically linked
functions), the use of our patch object format (discussed in detail in
Section \ref{sec:poformat}), and data patching including complex
structures and variable addition. The remainder of the system is still
under development. In the future we
%JMHO end
must address several 
important engineering issues, such as interaction of patched code with
dynamically loaded libraries (including the {\em dlopen} mechanism),
and assuring that accumulation of administered patches does not lead to
unacceptable performance degradation. It also poses a broader question
of describing and detecting software designs not amenable to its runtime
patching and steering programmers to avoid them if possible.

>>>>

{\bf When to apply the patch.}
%
Dynamically updating a running application requires diligence and
patience. One cannot update the target application without any
knowledge of the program's execution state, by which we mean the
program stack, processor registers, {\it etc}.  Even after possessing
this information, the application has to be in what we call a ``safe
state'' for Katana to apply the patch. We characterize a program state
as a {\it safe state} if the following two conditions hold:
\begin{itemize}
\item All activation frames in the program stack belong to functions
  that {\it do not} get updated during code patching. It is easy to
  verify this by comparing each function on the stack with the list of
  upgradeable functions contained within the PO.

\item All activation frames in the program stack belong to functions
  that do not access any global/static symbols identified during {\it
    Data Traversal} and that do not define any local variables of the
  modified types identified during {\it Type Discovery}. Again, since
  we maintain type and variable definitions within the PO, verifying
  this condition is easy.
\end{itemize}

Given the patch object, Katana uses Linux's {\tt ptrace} interface to
temporarily halt the execution of the target process, query the
current execution stack, and determine whether the application is currently
in a safe state. If so, then Katana applies the code patch followed by
the data patch. We note that it is not possible to apply the code and
data patches at different times since new code likely uses the new
data, and hence postponing data patching to when only the second
condition is satisfied is impractical and unsafe.

Let $A$ denote the current activation frame on the program stack,
and the notation $(X:Y)$ define all frames from $X$ to $Y$ (on a time
scale, $X$ precedes $Y$) on the stack; so $(1:A)$ defines the current
stack. Now, in case we determine $(1:A)$ to be an unsafe state, we
could repeatedly keep querying the stack until the application reaches
some safe state. However, this is highly inefficient and cumbersome.

@@

Finally we note that even after inserting a breakpoint, the problem of
determining {\it when} the breakpoint will be hit is essentially a
hard one, and so in such cases, Katana can provide no time-bounding
guarantees. Still, this is a cleaner and more efficient approach than
just na\"{i}vely retrying the full update procedure, which is both an
expensive and an incomplete solution.
%JMHO:
In cases where the timely (or even eventual) application of a patch
seems unlikely, it would be possible to add support for programmers to
manually specify safe places for patching to occur. For programs built
around a central loop (the vast majority of long-running programs),
the likelihood of safe patching can be expedited by isolating the main
loop into a simple routine unlikely ever to need patching. While
safety determination is implemented in our prototype, we do not yet
have enough real data about the likelihood of applications' failing to
reach a safe state.
%JHMO end

@@

Patching %``randomly'' 
relocated code with our PO files requires
knowledge of the displacements introduced at loading-and-relocation
time. While there is no common ABI standard for saving this
information, conceptually it is no different from saving virtual
addresses of other files' loaded symbols in the {\it Global Offset Table}
(GOT). We note that the names of the constituent object files themselves 
are customarily included in the symbol tables and that the symbol
table entry format can be easily adopted for storing virtual addresses
of the %``randomly'' 
relocated objects.

Thus, at the cost of small modifications to the OS loader and the 
dynamic linker, we can make the information on the layout of the
``randomly'' relocated executable and libraries available to
our patching process driven by our POs.\footnote{We note that saving 
this information 
%
about the post-relocation layout of the process 
does not weaken ``randomization,'' for 
the latter does not assume the attacker's ability to arbitrarily read 
process memory (in which case the addresses of required symbols are
easily found by scanning it %). %
for their code or data patterns), 
but rather breaks hard-coding of these symbols' expected addresses.
}

{\bf Future work.} Katana is a work in progress.
%JMHO:
We have demonstrated code patching (including dynamically linked
functions), the use of our patch object format (discussed in detail in
Section \ref{sec:poformat}), and data patching including complex
structures and variable addition. The remainder of the system is still
under development. In the future we
%JMHO end
must address several 
important engineering issues, such as the interaction of patched code with
dynamically loaded libraries (including the {\em dlopen} mechanism)
and assuring that accumulation of administered patches does not lead to
unacceptable performance degradation. 
%JMHO: reword
%% It also poses a broader question
%% of describing and detecting software designs not amenable to its runtime
%% patching and steering programmers to avoid them if possible. 
We must also address the broader issue of describing and detecting
software designs not amenable to runtime patching, and we must steer
programmers to avoid them if possible.

### 2010-03-27 paper1/hotpatch/likelinking.tex
made figure one smaller
removed numbered subsection headings, as required by style for this publication
many grammar fixes on advice from a retired english teacher

\subsection{Patching In The Toolchain}
\label{ssec:patchingvlinking}
Hot patching should not be thought of as a bizarre operation, done
crudely and infrequently, something akin to reverse engineering. We
argue that it is one of the fundamental transformations in the life
cycle of any program, along with compilation, linking, dynamic
linking, and (in unfortunate cases) dumping core. We note that each of
these fundamental operations has its own type of ELF object devoted to
it (relocatable objects, executables, dynamic libraries, core dumps
respectively) and a corresponding tool in the toolchain for effecting
each transformation or working with its output. \emph{We contend that
  patching is very much like linking or dynamic linking}. Like those
operations, it combines (or replaces) parts of programs and must
generate, modify, and apply relocation information. The section types
defined for the ELF format contain nearly all of the information
necessary to describe a patch. Once we take the position that patching
is like linking, a patch needs to store the same type of symbol and
relocation information as does any relocatable ELF object.

What the base ELF specification lacks is a way to describe
types. Symbol information gives us only a location and a length, but
there is no way to describe the internal layout of a piece of
data. The DWARF format, already heavily used with ELF for debugging
and exception handling purposes, provides exactly what is needed here
as it provides a means to recursively describe types and variables, as
well as a set of instructions originally designed for restoring
register states and examining the call stack but rich in possible
applications.

Through the use of formats already used in the binary tool chain, we
hope to promote easy examination of patches, interoperability, and to
show that patching fits comfortably in to the rest of the
toolchain. We therefore propose that the standard software life cycle
should now be as shown in Figure \ref{fig:lifecycle}. Before hot
patching, only the Development and Runtime stages of the figure were
generally accepted.

\begin{figure*}[ht]
\begin{center}
\includegraphics{software_lifecycle.pdf}
\end{center}
\caption{{\small Revised software life cycle. Before hot patching, only
    the Development and Runtime portions existed}}
\label{fig:lifecycle}
\end{figure*}

>>>>

%\subsection{Patching In The Toolchain}
{\bf Patching In The Toolchain.}
\label{ssec:patchingvlinking}
Hot patching should not be thought of as a bizarre operation, done
crudely and infrequently, something akin to reverse engineering. We
argue that it is one of the fundamental transformations in the life
cycle of any program, along with compilation, linking, dynamic
linking, and (in unfortunate cases) dumping core. We note that each of
these fundamental operations has its own type of ELF object devoted to
it (relocatable objects, executables, dynamic libraries, core dumps
respectively) and a corresponding tool in the toolchain for effecting
each transformation or working with its output. \emph{We contend that
patching is very much like linking or dynamic linking}. Like those
operations, it combines (or replaces) parts of programs and must
generate, modify, and apply relocation information. The section types
defined for the ELF format contain nearly all of the information
necessary to describe a patch. Once we take the position that patching
is like linking, it follows that a patch needs to store the same type
of symbol and relocation information as does any relocatable ELF
object.

What the base ELF specification lacks is a way to describe
types. Symbol information gives us only a location and a length, but
there is no way to describe the internal layout of a piece of
data. The DWARF format, already heavily used with ELF for debugging
and exception-handling purposes, provides exactly what is needed here
as it provides a means to recursively describe types and variables, as
well as a set of instructions originally designed for restoring
register states and examining the call stack but rich in possible
applications.

Through the use of formats already employed in the binary tool chain, we
hope to promote easy examination of patches, interoperability, and to
show that patching fits comfortably into the rest of the
toolchain. We therefore propose that the standard software life cycle
now be as shown in Figure \ref{fig:lifecycle}. Before hot
patching, only the Development and Runtime stages of the figure were
generally accepted.

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.5]{software_lifecycle.pdf}
\end{center}
\caption{{\small Revised software life cycle. Before hot patching, only
    the Development and Runtime portions existed}}
\label{fig:lifecycle}
\end{figure}
